---
title: "ESM244_lab2"
author: "Sam Csik"
date: "1/17/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. Attach packages
```{r}
library(tidyverse)
library(janitor)
library(MASS)
library(RColorBrewer)
library(factoextra)
library(ggbiplot)
library(shinythemes)
library(ggrepel)
```

If you want to override (set) a function manually thorughout an entire workspace, you can assign the function name explicitly:

```{r}
select <- dplyr::select
```

Or, include dplyr::select in line.

###1. Ordinal logistic regression for political party affiliation by ethnicity, age, sex

```{r}
pol_party <- read_csv("pol_party.csv") %>% 
  clean_names() %>% 
  select(participant_id:read_carefully)

# a little more wrangling
pol_df <- pol_party %>% 
  select(birth_year, gender, ethnicity, political_party) %>% 
  filter(birth_year > 1900, political_party != "NA", political_party <= 7) %>% 
  mutate(age = 2016 - birth_year)

# Political party: 1 = very liberal, 7 = very conservative

# When doing OLR we want our dependent variable to be of class ORDERED FACTOR

pol_df$political_party <- factor(pol_df$political_party, ordered = TRUE, levels = c("1", "2", "3", "4", "5", "6", "7"))

# Update our categorical predictor variables to factors too:

pol_df$ethnicity <- factor(pol_df$ethnicity)
pol_df$gender <- factor(pol_df$gender)
```

Basic data exploration and visualization:

```{r}
counts_eth <- pol_df %>% 
  group_by(ethnicity, political_party) %>% 
  tally()

counts_g <- pol_df %>% 
  group_by(gender, political_party) %>% 
  tally()

ggplot(pol_df, aes(x = political_party)) +
  geom_histogram(aes(fill = ethnicity), stat = "count") + 
  scale_fill_brewer(palette = "YlGnBu") + 
  facet_wrap(~ethnicity, scales = "free")
```

# THESE DATA ARE PRECARIOUS AND N IS TOO SMALL FOR GROUPS!! but let's see what happens anyway...

```{r}
# run OLR (will work as long as vars are of the correct class)

# THIS IS NOT PUBLISHABLE DATA B/C IT IS VERY BIASED

pol_model <- polr(political_party ~ age + gender + ethnicity, data = pol_df)
summary(pol_model)

# some interpretations
# as age increases, we would expect to see a TINY increase in the probability of someone ranking themselves...
# female was reference sex: increase in log odds relative to ref level; males are more likely ot rank higher on pol_preference scale relative to femaled

# since log odds are hard to think about, convert to odds by exponentiating
exp(coef(pol_model))

# odds = 1 means that they haven't changed from previous scenario
# age deosn't have an effect on pol_party affiliation
# increae in odds of males vs females of similar age and ethnicity

# Once we've decided on a model, we can always make predictions using it:

predictions <- predict(pol_model, type = "probs") # makes predictions for each person (by age, ethnicity) for where they will fall along the political_party affiliation

df <- data.frame(pol_df, predictions)
```

####2. Principal components analysis (PCA)

Data from the WorldBank environmental indicators

```{r}
wb_10 <- read_csv("wb_10.csv") %>% 
  select(-X1) %>% 
  column_to_rownames('code')

wb_pca <- prcomp(wb_10[3:8], scale = TRUE)

summary(wb_pca)

ggbiplot(wb_pca) + 
  geom_text_repel(label = rownames(wb_10), size = 2, segment.size = 0.2) +
  theme_bw()

# factoextra package (built to help visualize outcomes of ordination methods)

# screeplot:

fviz_screeplot(wb_pca)

# see contriburtions of variables to different components:

fviz_contrib(wb_pca, choice = "var", axes = 2, top = 6) # change axes to 1 or 2 to see contributers to each PCA

# Biplot showing points and vectors

fviz_pca_biplot(wb_pca,
                col.var = "cos2", 
                label = "var", 
                repel = TRUE)
  
```


